Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-03 Manuscript/TEM_draft.tex
Comments to methods up to tree comparisons

\section{Methods}
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}).
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the "true" tree) and used it to infer a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the "complete" matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the "complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). The resulting matrices are called hereafter "missing-data" matrices.
\item{Inferring phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the "complete" matrix and from the "missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the "best" tree) and multiple trees inferred from matrices with missing morphological data (hereafter called the "missing-data" trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the "best" tree to the "missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated the 4 steps 50 times to account for variation of our random parameters in the simulations.

\begin{figure}
\centering
    \includegraphics[keepaspectratio=true]{Figures/In_main/Simulations_outline-BW.pdf}
\caption{Protocol outline.
(1) We randomly generated a birth-death tree (the "true" tree) and used it to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological part of the "complete" matrix resulting in 125 "missing-data" matrices.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the "missing-data" trees to the "best" tree.
We repeated the 4 steps 50 times.}
\label{Fig_Outline}
\end{figure}

@@

\subsection{Generating the matrix}
First we randomly generated a "true" tree of 50 taxa in R v3.0.2 \citep{R302} using the package diversitree v0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. We implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa to ensure that we had enough taxa of each type for our missing data simulations to work. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the "true" tree. The molecular matrix was inferred from the "true" tree using the R package phyclust v0.1-14 \citep{chen2011}. The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies and transition/transversion rate of 2 \citep{douadycomparison2003}. The substitution rates were distributed following a gamma distribution with an alpha ($\alpha$) shape of 0.5 \citep{yangamong-site1996}. We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years). All the molecular information for fossil taxa was replaced by missing data ("?").

We inferred the morphological matrix using the R package ape v3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (See \hyperref[SupplementaryMaterial]{Supplementary Material Section 1}). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to an other \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014,wrightbayesian2014}.

Finally, we combined the morphological and molecular matrices obtained from the "true" tree. Hereafter we call this the "complete" matrix: the matrix with no missing data except for the molecular data of the fossil taxa.

@@

\begin{enumerate}
\item{The proportion of living taxa with no morphological data ($M_{L}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that because of the increasing availability of DNA sequences for living taxa, detailed morphological data is scarce.
\item{The proportion of missing data in the fossil taxa ($M_{F}$): 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record. 
\item{the proportion of missing morphological characters for both living and fossil taxa ($M_{C}$): 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa.
\end{enumerate}

In practice, each parameter represents a different way of removing data from the matrix: $M_{L}$ removes rows from the living taxa's data; $M_{F}$ removes cells from the fossil taxa's data; and $M_{C}$ removes columns across both living and fossil taxa's data. Note that $M_{L}$ is different to $M_{F}$ not only because of the region of the matrix affected: for $M_{L}$, all the morphological data of a percentage of living taxa is removed, but for $M_{F}$, a percentage of the data is removed at random from across the whole of the morphological matrix for fossil taxa.

We tested all parameters combinations resulting in 125 ($5^3$) matrices. Note that one of these combinations has no missing data so is equivalent to the "complete" matrix, thus we have one effectively complete matrix in our 125 "missing-data" matrices. Because some parameter combinations introduce a lot of missing data (e.g. $M_L$=75\%, $M_F$=75\% and $M_C$=75\%), some matrices contained fossil taxa without any data at all. When this occurred we repeated the random deletion of characters until every taxa had at least 5\% data across the whole morphological part of the matrix.

@@

\subsection{Building phylogenies}
From the resulting matrices we generated two types of trees, the "best" tree inferred from the "complete" matrix and the "missing-data" trees inferred from the 125 matrices with various amounts of missing data. The "true" tree was used to generate the "complete" matrix and reflects the "true" evolutionary history in our simulations. The "best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the "true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \citealt{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the "best" tree, rather than the "true" tree, as the "best" tree is generally what biologists have to work with.

\subsubsection{Maximum Likelihood}
The "best" tree and the "missing-data" trees were inferred using RAxML v8.0.20 \citep{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model (\citealt{tavare1986}; default GTRGAMMA in RAxML v8.0.20; \citealt{Stamatakis21012014})% as a generalization of the HKY + $\Gamma_4$ model \citep{HKY85}
for the molecular data.
%The GTR model can be seen as a generalization of the HKY model (the two parameters from the HKY model are implicitly included in the six from GTR model - \citealt{stamatakisa2008}). - TG: Is that useful? 
For the morphological data, we used the implemented Markov \textit{k} state model \citep{lewisa2001} %which is a generalization of the JC69 model \citep{jc69} with \textit{k} $\geq$ 2 - TG: same?
 assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v8.0.20; \citealt{Stamatakis21012014}).

In order to measure the phylogenetic signal of our simulations, we first ran a fast bootstrap analysis with 500 replicates on the "complete" matrix. We removed all the simulations that had a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50*125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50).

On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps per tree inference to assess the topological support \citep{pattengale2010many}.
%The bootstrap algorithm used in RAxML is the Lazy Sub-tree Rearrangement (LSR) which consists of pruning one sub-tree from the tree and subsequently reinserting it to all neighbouring branches \citep{stamatakisa2008}. Sub-tree Pruning and Reinserting methods (SPR) have been demonstrated to be better than other methods (e.g. Nearest Neighbouring Interchange - NNI) for recovering good bootstrap values \citep{salamin2003}. %TG: The explanations of the algorithm part for the BS was proposed by Trevor I think it's useless
When using these parameters, it took $~$8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

\subsubsection{Bayesian}
The "best" tree and the "missing-data" trees were inferred using MrBayes v3.2.1 \citep{Ronquist2012mrbayes}. We partitioned the data to treat the molecular part as a non-codon DNA partition and the morphological part as a multi-state morphological partition. The molecular evolutionary history was inferred using the HKY model with a transition/transversion ratio of 2 \citep{douadycomparison2003} and a gamma distribution for the rate variation with four distinct categories (HKY + $\Gamma_4$). For the morphological data, we used the Markov \textit{k} state model \citep{lewisa2001}, with equal state frequency and a unique overall substitution rate ($\mu$) with four distinct rates categories (M\textit{k} + $\Gamma_4$). We chose these models to be consistent with the parameters used to generate the "complete" matrix.

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 50$\times$$1^6$ generations. We used the average standard deviation of split frequencies (ASDS) as a proxy to estimate the convergence of the chains and used a stop rule when the ASDS went below 0.01 \citep{Ronquist2012mrbayes}. The effective sample size (ESS) was also checked on a random sub-sample of runs in each simulation to ensure that ESS $>>$ 200 \citep{drummond2006ess}. For each run, we removed 25\% of the iterations as burn-in. We used the following priors for each tree (see \hyperref[SupplementaryMaterial]{Supplementary Material S1}): (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of 1), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)).

We used these prior to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation, however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took $~$140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes).

@@

\subsection{Comparing topologies}
We compared the topology of the "missing-data" trees to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$ on tree topology. We used the Robinson-Foulds distance \citep{RF1981} to assess the amount conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess amount of wildcard taxa \citep{kearneyfragmentary2002}. We then used Normalized Tree Similarity index \citep{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa. These metrics are described in detail below.

\subsubsection{Robinson-Foulds distance}
Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981} ; see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). This metric is bounded between 1 when the two trees are identical and $n-2$ (for two trees with $n$ taxa) when there is not one single shared clade between both trees. This metric is sensitive to the exact clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swap of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating a bad tree similarity.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two given trees (\citealt{critchlowthe1996} ; see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). This metric measures the position of each taxon and clade towards it's closest neighbours. It is bounded between 0 when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is not one single position of taxon/clade identical between both trees. Therefore this metric sensitive to the conservation of wildcard taxas.

\subsubsection{Normalized Tree Similarity}
We used the Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) to be able to compare the two metrics for any $n$ taxa. This index allows to scale the value of any metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). When $NTS_m$=1, the two trees are strictly identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topology: a high score of this index (i.e. towards 1) means that the topology is highly conserved between the two trees; on the opposite, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as much as expected when comparing two random trees. Note that both tree comparisons are representing different aspects of topological similarity and do not perform equaly (see \hyperref[metrics_discussion]{discussion below} and \citealt{kuhnerpractical2014}).

\subsubsection{Tree comparisons}
\label{tree_comparisons}
For the Maximum Likelihood and Bayesian consensus trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Robinson-Foulds and Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012}. For each metric, we then normalized the value using the Normalized Tree Similarity scaled by the mean value of 1000 pairwise random tree comparisons for the metric in question and n = 51 taxa (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}). % TG: Maybe remove supplementary here
%We compared each "missing-data" tree with the "best" tree for each of our 50 simulation runs resulting in 50 comparisons for each "missing-data" tree. We calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}. TG: explained later on

Also, to take into account the uncertainty of tree inference in both Maximum Likelihood and Bayesian (i.e node support), we ran 1000 random pairwise comparison between respectively the bootstrapped trees from the Maximum Likelihood analysis and the posterior tree distribution of the Bayesian analysis. In the same way that we compared a single "missing-data" tree to the "best" tree (whether the trees are Maximum Likelihood or Bayesian consensus): we randomly selected 1000 trees from the "missing-data" tree sets (either the Bootstrapped trees or the posterior tree distribution) and did a pairwise comparison with 1000 randomly selected trees from the "best" tree set.

For each of the 125 "missing-data" tree, we obtained Robinson-Foulds and Triplets distance distributions from either 50 or 50$\times$1000 pairwise comparisons. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}. %Finally we tested for significant differences among the different distributions of trees using non-parametric group and pairwise comparisons by using respectively Kruskal-Wallis and Nemenyi-Damico-Wolfe-Dunn test (CITE R PACKAGE)\citet{ruxtontime2008}. TG: finally we didn't (Andrew's advice - see Battacharyya)

In order to investigate the effect of the parameters and/or the methods used in this simulations, we measured the similarity among the different distributions using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions, ranging from 0 to 1 (\citealt{Bhattacharyya} - see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}). For each method (Maximym likelihood trees, Bayesian consensus trees, Bootstraps and Bayesian posterior trees) and metric (Robinson-Foulds and Triplets distance), we used this coefficient in two ways in order to:
\begin{enumerate}
\item{Assess the effect of the method on all parameters:}

>>>>

\section{Methods}
To explore how missing data in the morphological sections of Total Evidence matrices influences tree topology, we used the following protocol (note that we explain each step in detail below this general outline; Fig.~\ref{Fig_Outline}).
% NC: I've added numbers in here (125 matrices etc) so it matches the figure legend better. Can you check I've put the right numbers in? I think it may be 124 in some cases...
\begin{enumerate}
\item{Generating the matrix:} \label{step:generate_matrix} \\
We randomly generated a birth-death tree (hereafter called the "true" tree) and used it to infer a matrix containing both molecular and morphological data for living and fossil taxa (hereafter called the "complete" matrix).
\item{Removing data:} \label{step:remove_data} \\
We removed data from the morphological part of the "complete" matrix to simulate the effects of missing data by modifying three parameters (i) the proportion of living taxa with no morphological data ($M_{L}$), (ii) the proportion of missing data in the fossil taxa ($M_{F}$) and (iii) the proportion of missing morphological characters ($M_{C}$). The resulting 125 matrices are called hereafter "missing-data" matrices.
\item{Inferring phylogenies:} \label{step:build_phylo} \\
We inferred phylogenetic trees from the "complete" matrix and from the 125 "missing-data" matrices resulting in one tree generated from a matrix containing no missing data (hereafter called the "best" tree) and 125 trees inferred from the matrices with missing morphological data (hereafter called the "missing-data" trees). Phylogenies were inferred via both Maximum Likelihood and Bayesian approaches.
\item{Comparing topologies:} \label{step:compare_topo} \\
We compared the "best" tree to the "missing-data" trees to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$) and their interactions on the topologies of our phylogenies
\end{enumerate}
We repeated these four steps 50 times to account for variation in our random parameters in the simulations.

\begin{figure}
\centering
    \includegraphics[keepaspectratio=true]{Figures/In_main/Simulations_outline-BW.pdf}
\caption{Protocol outline.
(1) We randomly generated a birth-death tree (the "true" tree) and used it to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological part of the "complete" matrix resulting in 125 "missing-data" matrices.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the "missing-data" trees to the "best" tree.
We repeated these four steps 50 times.}
\label{Fig_Outline}
\end{figure}
% NC: Can you centre these around the boxes? and make the comparing topologies arrow touch the two triangles? I'll show you what I mean. 
% NC: Also check the figure is in a format and of correct quality for Syst Biol.

@@

\subsection{Generating the matrix} % NC: Can you number these subsections like they are in the steps above?
First we randomly generated a "true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. We implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa to ensure that we had enough taxa of each type for our missing data simulations to work. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the "true" tree. The molecular matrix was inferred from the "true" tree using the R package phyclust v. 0.1-14 \citep{chen2011}. The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \citep{ranbaut1997seqgen} and using the HKY model \citep{HKY85} with random base frequencies and transition/transversion rate of two \citep{douadycomparison2003}. The substitution rates were distributed following a gamma distribution with an alpha ($\alpha$) shape of 0.5 \citep{yangamong-site1996}. We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years). All the molecular information for fossil taxa was replaced by missing data ("?").

We inferred the morphological matrix using the R package ape v. 3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (See \hyperref[SupplementaryMaterial]{Supplementary Material Section 1}). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014,wrightbayesian2014}.

Finally, we combined the morphological and molecular matrices obtained from the "true" tree. Hereafter we call this the "complete" matrix, i.e. the matrix with no missing data except for the molecular data of the fossil taxa.

@@

\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that detailed morphological data for living species are scarce.
% NC: Not sure you can really "blame" this on DNA data availability!
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record. 
\item{$M_{C}$, the proportion of missing morphological characters for both living and fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa.
\end{enumerate}

In practice, each parameter represents a different way of removing data from the matrix: $M_{L}$ removes rows from the living taxa's data; $M_{F}$ removes cells from the fossil taxa's data; and $M_{C}$ removes columns across both living and fossil taxa's data. Note that $M_{L}$ and $M_{F}$ differ not only because of the region of the matrix affected: for $M_{L}$ all the morphological data of a percentage of living taxa is removed, whereas for $M_{F}$ a percentage of the data is removed at random from across the whole of the morphological matrix for fossil taxa.

We created matrices using all parameter combinations resulting in 125 ($5^3$) matrices. Note that one of these combinations has no missing data so is equivalent to the "complete" matrix, thus we have one effectively complete matrix in our 125 "missing-data" matrices. Because some parameter combinations introduce a lot of missing data (e.g. $M_L$=75\%, $M_F$=75\% and $M_C$=75\%), some matrices contained fossil taxa without any data at all. When this occurred we repeated the random deletion of characters until each taxon had at least 5\% data across the whole morphological part of the matrix.

@@

\subsection{Building phylogenies}
From the resulting matrices we generated two types of trees: the "best" tree inferred from the "complete" matrix and the "missing-data" trees inferred from the 125 matrices with various amounts of missing data. The "true" tree was used to generate the "complete" matrix and reflects the "true" evolutionary history in our simulations. The "best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the "true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \citealt{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the "best" tree, rather than the "true" tree, as the "best" tree is generally what we have to work with.

\subsubsection{Maximum Likelihood}
The "best" tree and the "missing-data" trees were inferred using RAxML v. 8.0.20 \citep{Stamatakis21012014}. For the molecular data, we used the GTR + $\Gamma_4$ model (\citealt{tavare1986}; default GTRGAMMA in RAxML v. 8.0.20; \citealt{Stamatakis21012014}).% as a generalization of the HKY + $\Gamma_4$ model \citep{HKY85}.
%The GTR model can be seen as a generalization of the HKY model (the two parameters from the HKY model are implicitly included in the six from GTR model - \citealt{stamatakisa2008}). - TG: Is that useful? 
% NC: I think most people know that already.
For the morphological data, we used the implemented Markov \textit{k} state model \citep{lewisa2001} %which is a generalization of the JC69 model \citep{jc69} with \textit{k} $\geq$ 2 - TG: same?
% NC: Again leave it out
 assuming an equal state frequency and a unique overall substitution rate ($\mu$) following a gamma distribution of the rate variation with four distinct categories (M\textit{k} + $\Gamma_4$; -K MK option in RAxML v. 8.0.20; \citealt{Stamatakis21012014}).

To measure the phylogenetic signal of our simulations, we first ran a fast bootstrap analysis with 500 replicates on the "complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50 x 125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}.
%The bootstrap algorithm used in RAxML is the Lazy Sub-tree Rearrangement (LSR) which consists of pruning one sub-tree from the tree and subsequently reinserting it to all neighbouring branches \citep{stamatakisa2008}. Sub-tree Pruning and Reinserting methods (SPR) have been demonstrated to be better than other methods (e.g. Nearest Neighbouring Interchange - NNI) for recovering good bootstrap values \citep{salamin2003}. %TG: The explanations of the algorithm part for the BS was proposed by Trevor I think it's useless
% NC: perhaps somewhere in brackets just mention it's LSR bootstrapping. No need for the rest I agree.
Using these parameters took $~$8 % NC: again tilde isn't appearing on my PDF 
CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

\subsubsection{Bayesian}
The "best" tree and the "missing-data" trees were inferred using MrBayes v. 3.2.1 \citep{Ronquist2012mrbayes}. We partitioned the data to treat the molecular part as a non-codon DNA partition and the morphological part as a multi-state morphological partition. The molecular evolutionary history was inferred using the HKY model with a transition/transversion ratio of two \citep{douadycomparison2003} and a gamma distribution for the rate variation with four distinct categories (HKY + $\Gamma_4$). For the morphological data, we used the Markov \textit{k} state model \citep{lewisa2001}, with equal state frequency and a unique overall substitution rate ($\mu$) with four distinct rates categories (M\textit{k} + $\Gamma_4$). We chose these models to be consistent with the parameters used to generate the "complete" matrix.

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 50$\times$$10^6$ generations.
% NC: Do you mean 50 x 10^6? or 5 x 10^7? It's standard to use 10^x not 1^x
 We used the average standard deviation of split frequencies (ASDS) as a proxy to estimate the convergence of the chains and used a stop rule when the ASDS went below 0.01 \citep{Ronquist2012mrbayes}. The effective sample size (ESS) was also checked on a random sub-sample of runs in each simulation to ensure that ESS $>>$ 200 \citep{drummond2006ess}. For each run, we removed 25\% of the iterations as burn-in. We used the following priors for each tree (see \hyperref[SupplementaryMaterial]{Supplementary Material S1}
 % NC: For what? What's in the SI?
 ): (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of one), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)).

We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took $~$ % NC: Another tilde issue
140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes).

@@

\subsection{Comparing topologies}
We compared the topology of the "missing-data" trees to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$ on tree topology. We used the Robinson-Foulds distance \citep{RF1981} to assess the amount of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the amount of wildcard taxa \citep{kearneyfragmentary2002}. We then used the Normalized Tree Similarity index \citep{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa. These metrics are described in detail below.

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2}
% NC: For what? What's in the SI?
). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to even minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees (\citealt{critchlowthe1996}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2}
% NC: For what? What's in the SI?
). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between 0 when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxon/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa, i.e. taxa that frequently change position in different trees.
% NC: You may have a better definition of wildcard taxa than mine :)

\subsubsection{Normalized Tree Similarity}
The Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) allows us to compare the two metrics above for any number of taxa, by scaling the value of the metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}
% NC: Why? For more details? Simulations etc
). When $NTS_m$=1, the two trees are identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees, and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topological similarity: a high score of this index (i.e. towards one) means that the topology is highly conserved between the two trees; conversely, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as great as expected when comparing two random trees. Note that the two tree comparison metrics represent different aspects of topological similarity and do not perform equally well (see \hyperref[metrics_discussion]{discussion below} and \citealt{kuhnerpractical2014}).

% NC: If I understand right, you use RF normalised and TD normalised. NOT RF, TD and NTS. So having these as three separate sections seems odd. Instead I'd write RF section, then explain that it is not comparable across trees with different numbers of taxa so you normalise it as suggested in Bogdanowicz's NTS method. Then describe TD and explain that you normalise it as discussed above for RF. NTS isn't really a metric, just a simple normalisation procedure.

\subsubsection{Tree comparisons} % NC: Lost!!!!
\label{tree_comparisons}
For the Maximum Likelihood and Bayesian consensus trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Robinson-Foulds and Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012}. For each metric, we then normalized the value using the Normalized Tree Similarity scaled by the mean value of 1000 pairwise random tree comparisons for the metric in question and n = 51 taxa (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}). % TG: Maybe remove supplementary here
% NC: Yeah probably. You could also fit this into the section above. Again NTS isn't really a metric in my opinion...
%We compared each "missing-data" tree with the "best" tree for each of our 50 simulation runs resulting in 50 comparisons for each "missing-data" tree. We calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}. TG: explained later on

To take into account the uncertainty of tree inference in both Maximum Likelihood and Bayesian trees, we ran 1000 random pairwise comparisons among the bootstrapped trees from the Maximum Likelihood analysis and the posterior tree distribution of the Bayesian analysis. In the same way that we compared a single "missing-data" tree to the "best" tree (whether the trees are Maximum Likelihood or Bayesian consensus): we randomly selected 1000 trees from the "missing-data" tree sets (either the Bootstrapped trees or the posterior tree distribution) and did a pairwise comparison with 1000 randomly selected trees from the "best" tree set.

For each of the 125 "missing-data" trees, we obtained Robinson-Foulds and Triplets distance distributions from either 50 or 50$\times$1000 pairwise comparisons. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v. 3.1 \citep{hdrcde}. %Finally we tested for significant differences among the different distributions of trees using non-parametric group and pairwise comparisons by using respectively Kruskal-Wallis and Nemenyi-Damico-Wolfe-Dunn test (CITE R PACKAGE)\citet{ruxtontime2008}. TG: finally we didn't (Andrew's advice - see Battacharyya)

To investigate the effect of the parameters and/or the methods used in these simulations, we measured the similarity among the different distributions using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions, ranging from 0 to 1 (\citealt{Bhattacharyya} - see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}). For each method (Maximym likelihood trees, Bayesian consensus trees, Bootstraps and Bayesian posterior trees) and metric (Robinson-Foulds and Triplets distance), we used this coefficient in two ways in order to:
\begin{enumerate}
\item{Assess the effect of the method on all parameters:}
